---
title: "Week 11: Fairness vs. Social Welfare"
subtitle: "*DSAN 5450: Data Ethics and Policy*<br><span class='subsubtitle'>Spring 2024, Georgetown University</span>"
author: "Jeff Jacobs"
institute: "<a href='mailto:jj1088@georgetown.edu' target='_blank'>`jj1088@georgetown.edu`</a>"
bibliography: "../_DSAN5450.bib"
date: 2024-04-03
date-format: full
lecnum: 11
categories:
  - "Class Sessions"
format:
  revealjs:
    df-print: kable
    footer: "DSAN 5450 Week 11: {{< var w11.footer >}}"
    output-file: "slides.html"
    html-math-method: mathjax
    theme: ["../_jjslides.scss"]
    include-in-header:
      text: "<link href='https://cdn.jsdelivr.net/npm/bootstrap@5.3.2/dist/css/bootstrap.min.css' rel='stylesheet' integrity='sha384-T3c6CoIi6uLrA9TneNEoa7RxnatzjcDSCmG1MXxSR1GAsXEV/Dwwykc2MPK8M2HN' crossorigin='anonymous'><link rel='stylesheet' href='https://cdn.jsdelivr.net/npm/bootstrap-icons@1.11.3/font/bootstrap-icons.min.css'>"
    scrollable: true
    slide-number: true
    simplemenu:
      flat: true
      barhtml:
        header: "<div class='menubar'><span style='position: absolute; left: 8; padding-left: 8px;'><a href='./index.html'>&larr; Return to Notes</a></span><ul class='menu'></ul></div>"
      scale: 0.5
    revealjs-plugins:
      - simplemenu
  html:
    df-print: kable
    output-file: "index.html"
    html-math-method: mathjax
---

::: {.content-visible unless-format="revealjs"}

<center>
<a class="h2" href="./slides.html" target="_blank">Open slides in new window &rarr;</a>
</center>

:::

# Utility $\rightarrow$ Social Welfare {data-stack-name="Utility and Social Welfare"}

## Externalities {.crunch-title .smaller .crunch-quarto-figure}

* **Jeef** and **Keef** are roommates: Jeef loves listening to <a href='https://www.youtube.com/watch?v=OlQTn7gI8cw' target='_blank'>Tony Danza Tapdance Extravaganza</a>, but Keef is normal and slowly dies inside with each additional song

::: {.columns}
::: {.column width="50%"}

```{r}
#| label: externalities
library(tidyverse)
source("../_globals.r")
music_df <- tribble(
  ~Songs, ~Jeef, ~Keef,
  0, 0, 0,
  1, 13, -2,
  2, 18, -6,
  3, 24, -13,
  4, 28, -20,
  5, 30, -42
)
music_df <- music_df |>
  mutate(Total = Jeef + Keef)
music_df
```

:::
::: {.column width="50%"}

```{r}
#| label: roommate-plot
#| fig-height: 4.5
long_df <- music_df |>
  pivot_longer(!Songs, names_to="Roommate", values_to="Utility")
util_df <- long_df |>
  filter(Roommate != "Total")
ggplot(util_df, aes(x=Songs, y=Utility, color=Roommate)) +
  geom_line(linewidth=g_linewidth) +
  geom_point(size=g_pointsize) +
  labs(
    title="Individual Utility: Jeef vs. Keef",
    x="Number of Songs Played",
    y="Utility"
  ) +
  dsan_theme("quarter")
```

```{r}
#| label: welfare-plot
#| fig-height: 4.5
welfare_df <- long_df |>
  filter(Roommate == "Total")
ggplot(welfare_df, aes(x=Songs, y=Utility, color=Roommate)) +
  geom_line(linewidth=g_linewidth) +
  geom_point(size=g_pointsize) +
  labs(
    title="Social Welfare: Jeef and Keef",
    x="Number of Songs Played",
    y="Social Welfare"
  ) +
  scale_color_manual(values=c(cbPalette[3]), labels=c("Total      ")) +
  dsan_theme("quarter") +
  remove_legend_title()
```

:::
:::

## So What's the Issue?

* These utility values are **not observed**
* If we try to **elicit** them, both Jeef and Keef have **strategic incentives** to **lie** (over-exaggerate)
* Jeef maximizes own utility by reporting $u_j(s) = \infty$
* Keef maximizes own utility by reporting $u_k(s) = -\infty$
* (...Second price auctions tho)

## Now with Scarce Resources {.crunch-title .crunch-ul .crunch-math}

* In a given week, Jeef and Keef have **14 meals** and **7 aux hours** to divide amongst them

$$
\begin{align*}
\max_{m_1,m_2,a_1,a_2}& W(u_1(m_1,a_1),u_2(m_2,a_2)) \\
\text{s.t. }& m_1 + m_2 \leq 14 \\
\phantom{\text{s.t. }} & ~ \, a_1 + a_2 \; \leq 7
\end{align*}
$$

* Let's assume $u_i(m_i, a_i) = m_i + a_i$ for both
* $\Rightarrow$ One solution: $m_1 = 14, m_2 = 0, a_1 = 7, a_2 = 0$...
* $\Rightarrow$ Another: $m_1 = 0, m_2 = 14, a_1 = 0, a_2 = 7$...
* Who decides? Any decision implies $\omega_1, \omega_2$ ($\omega_1 + \omega_2 = 1$)

# The Dark Secret Behind Fairness in AI {data-stack-name="Fairness vs. Welfare"}

## The Conveniently-Left-Out Detail {.crunch-title .crunch-ul}

* Recall **predictive parity**:

$$
\mathbb{E}[Y \mid D = 1, A = 1] = \mathbb{E}[Y \mid D = 1, A = 0]
$$

* Who decides which $Y$ to pick?
* Answer: Whoever picks the **objective function**!
* For **profit-maximizing firm**: $\mathbb{E}[D (Y - c)]$
* For **welfare-maximizing society**: $W(u_1(D), \ldots, u_n(D))$
* Do these align? Sometimes yes, often no (affirmative action!)

## Remaining (Challenging) Details

* Who gets included in the SWF?
* People in one community? One state? One country?
* People in the future?
* Animals?

# Let's Talk Projects!

